On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-09-21 Manuscript/manuscript.tex
Responded to many of Prof Schifanos commets but table 2 needs more work and explanation

There are two types of data that are explored throughout this paper.  For the
signed-rank tests described in Section~\ref{sec:rank}, the data comes from 
athletes who competed in the 2022 World Championships and another competition 
(2019 World Championships, 2023 World Championships, or 2022 national 
championships). For the GLMM described in Section~\ref{sec:glmm}, data comes
from every world championships from 1999 to 2023. This investigation started 
shortly after the 2022 World Championships, but as the 2023 data has now become 
available, we are pleased to include it in both of our analyses, and it did not
significantly change our results.
\eds{I think you should somehow reference the source of the data here; I saw the
comment below about removing the Appendix and including the sources in the Data 
README file, so assuming the readers will have access to that file, you can 
simply refer to the sources within that file here. I will defer to Prof. Yan's
opinion, however!}

@@

Figure~\ref{fig:RankScatterplots} shows the data that will be analyzed
with the rank-based methods (described in Section~\ref{sec:rank} with results in 
Section~\ref{subsec:Results_Rank}). In each of the three graphs, the reaction 
times for athletes who competed
in the 2022 finals \eds{World Championships?} are compared to their reaction
times at another at which they competed.  
It is worth noting that each instance of the 2022 data is different, because
those who competed at both the 2019 and 2022 World Championships will be
different athletes than those who competed at both the 2022 and 2023 World
Championships. This data was only for athletes who competed
in at least one race at each championship, however it is possible for athletes
to have competed in one race in 2019 and two races in 2022. 
%\eds{did we not already cover the last two sentences in the previous paragraph, 
%or are you trying to say something different here?} \of{yes you are correct. I
%commented out the other one because I think this one flows better.}  
The 2022 World
Championship Heats and Semifinals both had on average lower reaction times than
their respective 2019 races, and the 2022 finals were substantially faster. It
is also worth noting is that the fastest reaction time in the 2022 World
Championships Finals and Semifinals both came from Devon Allen, but the 
difference between 0.101 and 0.099
seconds was the difference in what caused him to be disqualified.

@@

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the we chose a gamma mixed-effect model as our 
final model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. Two random effects were found useful: one capturing the venue effect in
each year and the other capturing the heat effect within each year.  The heat
effect is important as it captures the variability in the amount of time
athletes are on the starting blocks before the gun goes off.  Thus, we can try
to reduce this variability by accounting for it in the model through the heat
effect.
Let $Y_{ijk}$ be the reaction time of observation~$k$ in heat~$j$ in year~$i$.
Conditional on a venue effect $v_i$ of year~$i$ and a heat effect
$h_{i/j}$ nested within each year~$i$, $Y_{ijk}$ is modeled by a
gamma distribution. 
%\eds{I was talking to my husband about this. Could the heat effect be justified 
%to capture the variability in the amount of time between the ``set'' and the 
%gun going off (assuming this is different from heat to heat)? E.g., quicker 
%gunshots after the ``set'' could lead to slower reaction times (or the reverse), 
%but this effect would be shared across all athletes in the same heat.}
%\of{Yes I agree with this and updated the language in the text.}
In the classic generalized linear model (GLM) sense,
$Y_{ijk}$ has conditional mean $E(Y_{ijk} | v_i, h_{i/j}) > 0$ and a
dispersion parameter~$\phi > 0$. The conditional mean is characterized by
\[
g\{E(Y_{ijk} | v_i, h_{i/j})\} = \alpha + v_i + h_{i/j},
\]
where $g$ is a known link function,
$v_i$ is normally distributed with mean zero and
variance~$\sigma_v^2$, and $\delta_{i/j}$ 
\eds{$h_{i/j}$ or $\delta_{i/j}$? Table 2 uses $\delta_{i/j}$ (are you reporting
the BLUP for this? If yes, why not also report the BLUP of $v_i$ in Table 2?)}
is normally distributed with mean
zero and variance~$\sigma_h^2$.
Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time in cognitive
psychological research, the logarithm
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

@@

A summary of the two effect model is shown in Table~\ref{tab:Gamma_parameters}
for the pooled datasets and shows the standard error of the year
effect increases when including 2022.
\eds{this is not what is currently being shown in
Table~\ref{tab:Gamma_parameters}; also, consider changing ``Full Model'' to 
``Year and Heat Effects''}  
The GLMM with both venue
%two effect model with venue 
\eds{let's try to be consistent with using either
`year' or `venue' in the text, and not both, except when you introduce the intuition behind its use-- it doesn't make a difference to 
me which one you use} 
and heat random
effects was the best model as evidenced by a decrease in the AIC compared to the
other models with only one random effect. %For the entire data \eds{??}, the 
%AIC for the venue effect only model was
%$-2023.5$, the AIC for the heat effect only model $-2070.7$, and the AIC for the 
%two effect model, which we will present throughout this section, was $-2075.7$. 
Thus, results from the GLMM with both venue and heat random effects %is superior and 
will be presented throughout the rest
of the paper.
%\of{How specific do I need to be in this section? Will people who read this paper
%run the code to see the p value or should I say it directly and how do I say it?}
%\eds{If you are doing a likelihood ratio test, I suggest including the p-value 
%and the degrees of freedom associated with the test.}
%\of{Professor Yan recommended not even mentioning the likelihood ratio so
%instead I will just talk about the AIC.}

\begin{table}
  \centering
  \caption{Estimated parameters for the two effect GLMM on various data sets. SE
  is short for standard error. %\eds{data in table needs to be revisited and table
	%needs to be referenced within the text. Also, it needs to be clarified 
	%somewhere that ``Old Pooled'' excludes 2022 and ``All Pooled'' includes 2022.}
  %\of{Resolved}
	\eds{This table is not showing what you have described here or in the main text.}} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & AIC & $\sigma_h^2$ & $\sigma_v^2$ & $\alpha$ & $\delta_{i/j}$ \\
      \midrule
      Year Effect Only & $-2023.5$ & N/A & 0.048 & $-1.857$ & 0.152\\
      Heat Effect Only & $-2070.7$ & 0.073 & N/A & $-1.890$ & 0.135\\
      Full model       & $-2075.7$ & 0.057 & 0.040 & $-1.880$ & 0.134\\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

>>>>

There are two types of data that are explored throughout this paper.  For the
signed-rank tests described in Section~\ref{sec:rank}, the data comes from 
athletes who competed in the 2022 World Championships and another competition 
(2019 World Championships, 2023 World Championships, or 2022 national 
championships). For the GLMM described in Section~\ref{sec:glmm}, data comes
from every world championships from 1999 to 2023. This investigation started 
shortly after the 2022 World Championships, but as the 2023 data has now become 
available, we are pleased to include it in both of our analyses, and it did not
significantly change our results \citep{WAData}.
\eds{I think you should somehow reference the source of the data here; I saw the
comment below about removing the Appendix and including the sources in the Data 
README file, so assuming the readers will have access to that file, you can 
simply refer to the sources within that file here. I will defer to Prof. Yan's
opinion, however!}
\of{Is this what you meant?}

@@

Figure~\ref{fig:RankScatterplots} shows the data that will be analyzed
with the rank-based methods (described in Section~\ref{sec:rank} with results in 
Section~\ref{subsec:Results_Rank}). In each of the three graphs, the reaction 
times for athletes who competed at the 2022 World Championships 
\eds{World Championships?}
\of{Resolved} are compared to their reaction
times at another at which they competed.  
It is worth noting that each instance of the 2022 data is different, because
those who competed at both the 2019 and 2022 World Championships will be
different athletes than those who competed at both the 2022 and 2023 World
Championships. This data was only for athletes who competed
in at least one race at each championship, however it is possible for athletes
to have competed in one race in 2019 and two races in 2022. 
%\eds{did we not already cover the last two sentences in the previous paragraph, 
%or are you trying to say something different here?} \of{yes you are correct. I
%commented out the other one because I think this one flows better.}  
The 2022 World
Championship Heats and Semifinals both had on average lower reaction times than
their respective 2019 races, and the 2022 finals were substantially faster. It
is also worth noting is that the fastest reaction time in the 2022 World
Championships Finals and Semifinals both came from Devon Allen, but the 
difference between 0.101 and 0.099
seconds was the difference in what caused him to be disqualified.

@@

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the we chose a gamma mixed-effect model as our 
final model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. Two random effects were found useful: one capturing the venue effect in
each year and the other capturing the heat effect within each year.  The heat
effect is important as it captures the variability in the amount of time
athletes are on the starting blocks before the gun goes off.  Thus, we can try
to reduce this variability by accounting for it in the model through the heat
effect.
Let $Y_{ijk}$ be the reaction time of observation~$k$ in heat~$j$ in year~$i$.
Conditional on a venue effect $v_i$ of year~$i$ and a heat effect
$h_{i/j}$ nested within each year~$i$, $Y_{ijk}$ is modeled by a
gamma distribution. 
%\eds{I was talking to my husband about this. Could the heat effect be justified 
%to capture the variability in the amount of time between the ``set'' and the 
%gun going off (assuming this is different from heat to heat)? E.g., quicker 
%gunshots after the ``set'' could lead to slower reaction times (or the reverse), 
%but this effect would be shared across all athletes in the same heat.}
%\of{Yes I agree with this and updated the language in the text.}
In the classic generalized linear model (GLM) sense,
$Y_{ijk}$ has conditional mean $E(Y_{ijk} | v_i, h_{i/j}) > 0$ and a
dispersion parameter~$\phi > 0$. The conditional mean is characterized by
\[
g\{E(Y_{ijk} | v_i, h_{i/j})\} = \alpha + v_i + h_{i/j},
\]
where $g$ is a known link function,
$v_i$ is normally distributed with mean zero and
variance~$\sigma_v^2$, and $\delta_{i/j}$ 
\eds{$h_{i/j}$ or $\delta_{i/j}$? Table 2 uses $\delta_{i/j}$ (are you reporting
the BLUP for this? If yes, why not also report the BLUP of $v_i$ in Table 2?)}
\of{I think this should say $h_{i/j}$. In fact I am not sure we ever properly 
defined what $\delta_{i/j}$ is. The number in the last column of the table is
what shows up as the residual standard deviation in R}
is normally distributed with mean
zero and variance~$\sigma_h^2$.
Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time in cognitive
psychological research, the logarithm
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

@@

The estimated parameters of the GLMM model from our analysis are shown in 
Table~\ref{tab:Gamma_parameters}.  These statistics reflect the inclusion of 2022
in our analysis and shows the three models we considered. $\sigma_h^2$ is the
standard deviation of the heat effect, $\sigma_v^2$ is the standard deviation of
the venue effect, and $\alpha$ is the intercept.
\eds{this is not what is currently being shown in
Table~\ref{tab:Gamma_parameters}; also, consider changing ``Full Model'' to 
``Year and Heat Effects''}  
\of{Resolved}
The GLMM with both venue
%two effect model with venue 
\eds{let's try to be consistent with using either `year' or `venue' in the text,
and not both, except when you introduce the intuition behind its use-- it 
doesn't make a difference to me which one you use} 
\of{We have been using venue so I will use that, I think I made a typo while
editing the table.}
and heat random
effects was the best model as evidenced by a decrease in the AIC compared to the
other models with only one random effect. %For the entire data \eds{??}, the 
%AIC for the venue effect only model was
%$-2023.5$, the AIC for the heat effect only model $-2070.7$, and the AIC for the 
%two effect model, which we will present throughout this section, was $-2075.7$. 
Thus, results from the GLMM with both venue and heat random effects %is superior and 
will be presented throughout the rest
of the paper.
%\of{How specific do I need to be in this section? Will people who read this paper
%run the code to see the p value or should I say it directly and how do I say it?}
%\eds{If you are doing a likelihood ratio test, I suggest including the p-value 
%and the degrees of freedom associated with the test.}
%\of{Professor Yan recommended not even mentioning the likelihood ratio so
%instead I will just talk about the AIC.}

\begin{table}
  \centering
  \caption{Estimated parameters for the two effect GLMM on various data sets. SE
  is short for standard error. %\eds{data in table needs to be revisited and table
	%needs to be referenced within the text. Also, it needs to be clarified 
	%somewhere that ``Old Pooled'' excludes 2022 and ``All Pooled'' includes 2022.}
  %\of{Resolved}
	\eds{This table is not showing what you have described here or in the main text.}} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & AIC & $\sigma_h^2$ & $\sigma_v^2$ & $\alpha$ & Std Dev Residual \\
      \midrule
      Venue Effect Only & $-2023.5$ & N/A & 0.054 & $-1.871$ & 0.150\\
      Heat Effect Only & $-2070.7$ & 0.073 & N/A & $-1.890$ & 0.135\\
      Venue and Heat Effect & $-2075.7$ & 0.057 & 0.040 & $-1.880$ & 0.134\\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}
